Relative Stable Maps in Genus One via Central Alignments

### 2018-11-17 genus-one-relative-maps.tex

\usepackage[english]{babel}
\usepackage{appendix}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
%\usepackage{showlabels}
\usepackage{hyperref}
\usepackage{amsthm}
\usepackage{marginnote}
\usepackage{stmaryrd}
\usepackage{enumitem}
\usepackage[english]{babel}
\usepackage{yfonts}
\usepackage[T1]{fontenc}
\usepackage[utf8x]{inputenc}

@@

@@

\newcommand{\pM}{\pazocal{M}}
\newcommand{\TT}{\operatorname{T}}
\newcommand{\oM}{\overline{\mathcal{M}}}
\newcommand{\M}[4]{\overline{\mathcal{M}}_{#1,#2}(#3,#4)}
\newcommand{\Q}[4]{\mathcal{Q}_{#1,#2}(#3,#4)}
\newcommand{\Qe}[4]{\mathcal{Q}^{\epsilon}_{#1,#2}(#3,#4)}
\newcommand{\Qt}[4]{\widetilde{\mathcal Q}_{#1,#2}(#3,#4)}
\newcommand{\QG}[4]{\mathcal{Q}G_{#1,#2}(#3,#4)}
\newcommand{\QGe}[4]{\mathcal{Q}G^{\epsilon}_{#1,#2}(#3,#4)}
\newcommand{\D}[3]{\mathcal{D^Q}(#1,#2,#3)}
\newcommand{\E}[3]{\mathcal{E^Q}(#1,#2,#3)}
\newcommand{\PP}{\mathbb P}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\tVZc}[4]{\widetilde{\mathcal{V\!Z}}^{\rm{ctr}}_{#1,#2}(#3,#4)}
\newcommand{\VZc}[4]{\mathcal{V\!Z}^{\rm{ctr}}_{#1,#2}(#3,#4)}
\newcommand{\VZcLi}[4]{\mathcal{V\!Z}^{\rm{ctr, Li}}_{#1,#2}(#3,#4)}
\newcommand{\VZrel}[4]{\mathcal{V\!Z}^{\rm{rel}}_{#1,#2}(#3,#4)}
\newcommand{\stab}{\rm{stab}}
\newcommand{\stC}{C'}
\newcommand{\stf}{f'}
\newcommand{\stpi}{\pi'}
\newcommand{\sts}{s'}
\newcommand{\N}{\mathbb{N}}
\newcommand{\OO}{\mathcal{O}}
\renewcommand{\to}{\rightarrow}
\newcommand{\A}{\mathcal A}
\newcommand{\B}{\mathcal B}
\newcommand{\C}{\mathfrak C}
\newcommand{\cC}{\mathcal C}
\newcommand{\EE}{\mathbf{E}}
\renewcommand{\L}{\mathcal L}
\newcommand{\LL}{\mathbf{L}}
\newcommand{\MM}{\mathfrak M}
\newcommand{\Aaff}{\mathbb{A}}
\newcommand{\kfield}{\Bbbk}
\newcommand{\comp}{\chi}
\newcommand{\sst}{\sigma^{\operatorname{ss}}}
\newcommand{\Pic}{\operatorname{Pic}}
\newcommand{\Def}{\operatorname{Def}}
\newcommand{\Spec}{\operatorname{Spec}}
\newcommand{\Proj}{\operatorname{Proj}}
\newcommand{\Hom}{\operatorname{Hom}}
\newcommand{\Ext}{\operatorname{Ext}}
\newcommand{\Gm}{\mathbb{G}_{\text{m}}}
\newcommand{\virt}[1]{[#1]^{\operatorname{virt}}}
\newcommand{\vip}[1]{[#1]^{\operatorname{prod}}}
\newcommand{\Id}{\operatorname{Id}}
\newcommand{\CC}{\mathbb{C}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\HH}{\operatorname{H}}
\newcommand{\Achow}{\operatorname{A}}
\newcommand{\pt}{\operatorname{pt}}
\newcommand{\bq}{\begin{equation}}
\newcommand{\eq}{\end{equation}}
\newcommand{\ba}{\begin{aligned}}
\newcommand{\ea}{\end{aligned}}
\newcommand{\be}{\begin{enumerate}}
\newcommand{\ee}{\end{enumerate}}
\newcommand{\bsm}{\left(\begin{smallmatrix}}
\newcommand{\esm}{\end{smallmatrix}\right)}                   
\newcommand{\bpm}{\begin{pmatrix}}
\newcommand{\epm}{\end{pmatrix}}
\newcommand{\barr}{\begin{displaymath}\begin{array}{cccc}}
\newcommand{\earr}{\end{array}\end{displaymath}}
\newcommand{\barrl}{\begin{displaymath}\begin{array}{lcl}}
\newcommand{\earrl}{\end{array}\end{displaymath}}
\newcommand{\barl}{\begin{displaymath}\begin{array}{l}}
\newcommand{\earl}{\end{array}\end{displaymath}}
\newcommand{\bxym}{ \begin{displaymath}\xymatrix }
\newcommand{\exym}{\end{displaymath}}
\newcommand{\bcd}{\begin{center}\begin{tikzcd}}
\newcommand{\ecd}{\end{tikzcd}\end{center}}
\newcommand{\R}{\operatorname{R}^{\bullet}}
\newcommand{\dvr}{\Delta}
%\newcommand{\sslash}{\mathbin{/\mkern-6mu/}}
\newcommand{\tr}{{\rm tr}}
\newcommand{\Isom}{\text{Isom}}
\newcommand{\pr}{\operatorname{pr}}
\newcommand{\ev}{\operatorname{ev}}
\newcommand{\codim}{\operatorname{codim}}
\newcommand{\vdim}{\operatorname{vdim}}
\newcommand{\ildef}[1]{\emph{#1}}
\newcommand{\om}[1]{\mathcal{#1}}
\newcommand{\h}{\operatorname{h}}
\newcommand{\Aut}{\operatorname{Aut}}
\newcommand{\RR}{\textbf{R}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\ovm}[1]{\overline{\mathcal{#1}}}
\newcommand{\ovt}[1]{\widetilde{\mathcal{#1}}}
\newcommand{\ov}[1]{\overline{#1}}

@@

\section{The space of relative centrally aligned maps}
\noindent Recall \cite{RSPW} that the moduli space of maps from centrally aligned curves is obtained by considering the  Cartesian diagram
\bcd
\tVZc{1}{n}{X}{\beta}\ar[d]\ar[r]\ar[dr,phantom,"\Box"] & \M{1}{n}{X}{\beta}\ar[d] \\
\MM_{1,n}^{\rm{ctr}}\ar[r] & \MM_{1,n}^{\dagger}
\ecd
so that objects of $\tVZc{1}{n}{X}{\beta}$ consist of
\begin{enumerate}
\item a centrally aligned curve $(C,M_C,\delta)$;
\item a stable map $f\colon C \to X$;
\end{enumerate}
subject to the condition that the subcurve $C_0 \subseteq C$, consisting of those components $C_v$ for which $\lambda(v) < \delta$, coincides with the maximal connected genus one subcurve contracted by $f$. They then define
\begin{equation*} \VZc{1}{n}{X}{\beta} \subseteq \tVZc{1}{n}{X}{\beta} \end{equation*}
to be the closed substack consisting of maps satisfying the \emph{factorisation condition}, namely that the map $f\colon C\to X$ factors through the associated contraction to a Smyth curve, i.e. there exists a map $\bar{f}$ making the following square commute:
\bcd
\widetilde C\ar[r]\ar[d] & \overline C\ar[d,"\bar f" left,dotted] \\
C\ar[r,"f"] & X
\ecd
One should think of the factorisation condition as identifying the main component of the moduli space.

\section{Definition of centrally aligned relative space}
\section{Characterisation of the closed points of the relative space}
\section{Recursion formula}

>>>>

\usepackage[english]{babel}
\usepackage{appendix}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
%\usepackage{showlabels}
\usepackage[hidelinks]{hyperref}
\usepackage{amsthm}
\usepackage{marginnote}
\usepackage{stmaryrd}
\usepackage{enumitem} 
\usepackage[english]{babel}
\usepackage{yfonts}
\usepackage[T1]{fontenc}
\usepackage[utf8x]{inputenc}

@@

\newcommand{\basenote}[1]{\footnote{\color{red}#1}}

@@

\newcommand{\formLine}{\spec \mathbbm{k}\llbracket t \rrbracket}
\newcommand{\spec}{\operatorname{Spec}}
\newcommand{\pM}{\pazocal{M}}

\newcommand{\TT}{\operatorname{T}}
\newcommand{\oM}{\overline{\mathcal{M}}}
\newcommand{\M}[4]{\overline{\mathcal{M}}_{#1,#2}(#3,#4)}
\newcommand{\Q}[4]{\mathcal{Q}_{#1,#2}(#3,#4)}
\newcommand{\Qe}[4]{\mathcal{Q}^{\epsilon}_{#1,#2}(#3,#4)}
\newcommand{\Qt}[4]{\widetilde{\mathcal Q}_{#1,#2}(#3,#4)}
\newcommand{\QG}[4]{\mathcal{Q}G_{#1,#2}(#3,#4)}
\newcommand{\QGe}[4]{\mathcal{Q}G^{\epsilon}_{#1,#2}(#3,#4)}
\newcommand{\D}[3]{\mathcal{D^Q}(#1,#2,#3)}
\newcommand{\E}[3]{\mathcal{E^Q}(#1,#2,#3)}
\newcommand{\PP}{\mathbb P}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\tVZc}[4]{\widetilde{\mathcal{V\!Z}}^{\rm{ctr}}_{#1,#2}(#3,#4)}
\newcommand{\VZc}[4]{\mathcal{V\!Z}^{\rm{ctr}}_{#1,#2}(#3,#4)}
\newcommand{\VZclb}[4]{\mathcal{V\!Z}^{\rm{ctr,lb}}_{#1,#2}(#3,#4)}
\newcommand{\VZco}[4]{\mathcal{V\!Z}^{\rm{ctr},\circ}_{#1,#2}(#3,#4)}
\newcommand{\VZcLi}[4]{\mathcal{V\!Z}^{\rm{ctr, Li}}_{#1,#2}(#3,#4)}
\newcommand{\VZrel}[4]{\mathcal{V\!Z}^{\rm{rel}}_{#1,#2}(#3,#4)}
\newcommand{\stab}{\rm{stab}}
\newcommand{\stC}{C'}
\newcommand{\stf}{f'}
\newcommand{\stpi}{\pi'}
\newcommand{\sts}{s'}
\newcommand{\N}{\mathbb{N}}
\newcommand{\OO}{\mathcal{O}}
\renewcommand{\to}{\rightarrow}
\newcommand{\A}{\mathcal A}
\newcommand{\B}{\mathcal B}
\newcommand{\C}{\mathfrak C}
\newcommand{\cC}{\mathcal C}
\newcommand{\EE}{\mathbf{E}}
\renewcommand{\L}{\mathcal L}
\newcommand{\LL}{\mathbf{L}}
\newcommand{\MM}{\mathfrak M}
\newcommand{\Aaff}{\mathbb{A}}
\newcommand{\kfield}{\Bbbk}
\newcommand{\comp}{\chi}
\newcommand{\sst}{\sigma^{\operatorname{ss}}}
\newcommand{\Pic}{\operatorname{Pic}}
\newcommand{\Def}{\operatorname{Def}}
\newcommand{\Spec}{\operatorname{Spec}}
\newcommand{\Proj}{\operatorname{Proj}}
\newcommand{\Hom}{\operatorname{Hom}}
\newcommand{\Ext}{\operatorname{Ext}}
\newcommand{\Gm}{\mathbb{G}_{\text{m}}}
\newcommand{\virt}[1]{[#1]^{\operatorname{virt}}}
\newcommand{\vip}[1]{[#1]^{\operatorname{prod}}}
\newcommand{\Id}{\operatorname{Id}}
\newcommand{\CC}{\mathbb{C}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\HH}{\operatorname{H}}
\newcommand{\Achow}{\operatorname{A}}
\newcommand{\pt}{\operatorname{pt}}
\newcommand{\bq}{\begin{equation}}
\newcommand{\eq}{\end{equation}}
\newcommand{\ba}{\begin{aligned}}
\newcommand{\ea}{\end{aligned}}
\newcommand{\be}{\begin{enumerate}}
\newcommand{\ee}{\end{enumerate}}
\newcommand{\bsm}{\left(\begin{smallmatrix}}
\newcommand{\esm}{\end{smallmatrix}\right)}                   
\newcommand{\bpm}{\begin{pmatrix}}
\newcommand{\epm}{\end{pmatrix}}
\newcommand{\barr}{\begin{displaymath}\begin{array}{cccc}}
\newcommand{\earr}{\end{array}\end{displaymath}}
\newcommand{\barrl}{\begin{displaymath}\begin{array}{lcl}}
\newcommand{\earrl}{\end{array}\end{displaymath}}
\newcommand{\barl}{\begin{displaymath}\begin{array}{l}}
\newcommand{\earl}{\end{array}\end{displaymath}}
\newcommand{\bxym}{ \begin{displaymath}\xymatrix }
\newcommand{\exym}{\end{displaymath}}
\newcommand{\bcd}{\begin{center}\begin{tikzcd}}
\newcommand{\ecd}{\end{tikzcd}\end{center}}
\newcommand{\R}{\operatorname{R}^{\bullet}}
\newcommand{\dvr}{\Delta}
%\newcommand{\sslash}{\mathbin{/\mkern-6mu/}}
\newcommand{\tr}{{\rm tr}}
\newcommand{\Isom}{\text{Isom}}
\newcommand{\pr}{\operatorname{pr}}
\newcommand{\ev}{\operatorname{ev}}
\newcommand{\codim}{\operatorname{codim}}
\newcommand{\vdim}{\operatorname{vdim}}
\newcommand{\ildef}[1]{\emph{#1}}
\newcommand{\om}[1]{\mathcal{#1}}
\newcommand{\h}{\operatorname{h}}
\newcommand{\Aut}{\operatorname{Aut}}
\newcommand{\RR}{\textbf{R}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\ovm}[1]{\overline{\mathcal{#1}}}
\newcommand{\ovt}[1]{\widetilde{\mathcal{#1}}}
\newcommand{\ov}[1]{\overline{#1}}

\newcommand{\Ccal}{\mathcal{C}}
\newcommand{\Mcal}{\mathcal{M}}
\newcommand{\Lcal}{\mathcal{L}}
\newcommand{\Mfrak}{\mathfrak{M}}

@@

\section{The space of relative centrally aligned maps}
\noindent Recall \cite{RSPW} that the moduli space of maps from centrally aligned curves is obtained by first considering the Cartesian diagram
\bcd
\tVZc{1}{n}{X}{\beta}\ar[d]\ar[r]\ar[dr,phantom,"\Box"] & \M{1}{n}{X}{\beta}\ar[d] \\
\MM_{1,n}^{\rm{ctr}}\ar[r] & \MM_{1,n}^{\dagger}
\ecd
so that objects of $\tVZc{1}{n}{X}{\beta}$ consist of
\begin{enumerate}
\item a centrally aligned curve $(C,\Mcal_C,\delta)$;
\item a stable map $f\colon C \to X$;
\end{enumerate}
subject to the condition that the subcurve $C_0 \subseteq C$, consisting of those components $C_v$ for which $\lambda(v) < \delta$, coincides with the maximal connected genus one subcurve contracted by $f$. They then define
\begin{equation*} \VZc{1}{n}{X}{\beta} \subseteq \tVZc{1}{n}{X}{\beta} \end{equation*}
to be the closed substack consisting of maps satisfying the \emph{factorisation condition}, namely that the map $f\colon C\to X$ factors through the associated contraction to a Smyth curve, i.e. there exists a map $\bar{f}$ making the following square commute:
\bcd
\widetilde C\ar[r]\ar[d] & \overline C\ar[d,"\bar f" left,dotted] \\
C\ar[r,"f"] & X
\ecd
One should think of the factorisation condition as identifying the main component of the moduli space.

\section{Definition of centrally aligned relative space}
Fix a vector $\alpha=(\alpha_1,\ldots,\alpha_n)$ of tangency conditions, and suppose for the moment that $\Sigma \alpha = d$. We consider the so-called \emph{nice locus} of relative centrally aligned maps. This is the locally closed substack of $\VZc{1}{n}{\PP^N}{d}$ consisting of centrally aligned maps $(C,x_1,\ldots,x_n,\Mcal_C,\delta,f)$ such that:
\begin{enumerate}
\item the source curve $C$ is smooth;
\item $C$ is not mapped entirely into $H$ by $f$;
\item the points of $C$ which are mapped into $H$ (necessarily a finite number by the previous condition) are all marked, with the map having tangency $\alpha_i$ to $H$ at the marking $x_i$.
\end{enumerate}
These conditions mean that $\delta=0$ and that the log structure $\Mcal_C$ (which is a minimal centrally aligned log structure) is equal to the minimal structure for a log smooth curve, which in this case (since there are no nodes) is just the divisorial log structure with respect to the markings $x_1,\ldots,x_n$.

This is a locally closed substack of expected codimension $\Sigma \alpha=d$, and is denoted:
\begin{equation*} \VZco{1}{\alpha}{\PP^N|H}{d} \end{equation*}
It has a well-defined stack structure since it is a moduli space. Note that it is irreducible, since we can find a parametrisation for it over $\Mfrak_{1,n}$.

We now define the \emph{centrally aligned relative space} to be the closure of the nice locus:
\begin{equation*} \VZc{1}{\alpha}{\PP^N|H}{d} := \overline{\VZco{1}{\alpha}{\PP^N|H}{d}} \subseteq \VZc{1}{n}{\PP^N}{d} \end{equation*}
This inherits the natural structure of a stack from the stack structure of the nice locus. It is proper since it is a closed substack of a proper stack, and it is irreducible because it is the closure of an irreducible stack. Hence it admits a fundamental class, which we can use to define \emph{reduced relative Gromov--Witten invariants}:
\begin{equation*} \langle \tau_{k_1}(\gamma_1), \ldots, \tau_{k_n}(\gamma_n) \rangle^{\PP^N|H, \operatorname{red}}_{0,\alpha,d} := \int_{[\VZc{1}{\alpha}{\PP^N|H}{d}]} \prod_{i=1}^n \ev_i^*(\gamma_i) \cdot \psi_i^{k_i} \end{equation*}

\section{Characterisation of the closed points of the relative space}
It will be important for our purposes to have an understanding of the boundary of the relative space, i.e. what we add in when we take the closure of the nice locus. In this section we will describe this boundary on the level of closed points.

\begin{prop} \label{closed pts of relative space} A centrally aligned map over a closed point $\spec \mathbbm{k}$ is in the relative space if and only if the following two conditions are satisfied:
\begin{enumerate}
\item \emph{Gathmann's relative condition:} for each connected component $Z$ of $f^{-1}(Y)$, either:
\begin{enumerate}
\item $Z$ is a single point of $C$, in which case if $Z=x_i$ is a marking then we require that the multiplicity of $f$ at $x_i$ along $Y$ is at least $\alpha_i$ (if $Z$ is not a marking then there is no condition);
\item $Z$ is a subcurve $Z \subseteq C$, in which case we require that
\begin{equation*} f^*[Y]|_Z -\sum_{x_i\in Z}\alpha_ix_i\end{equation*}
is an effective class in $\Achow_0(Z)$.
\end{enumerate}

\item \emph{Compatibility condition:} this applies when the minimal subcurve of genus one (the \emph{circuit}) is contracted to a point in $Y$. In this case we require that there is a local morphism of monoids:
\begin{equation*} Q^{\operatorname{GS}} \to \N\end{equation*}
such that the dashed arrow can be filled in to make the following square commutative:
\bcd
Q^{\operatorname{log}} \ar[r]\ar[d] & Q^{\operatorname{GS}} \ar[d] \\
Q^{\operatorname{ctr}}\ar[r,dashed] & \N
\ecd
It is obvious from the construction that the map $Q^{\operatorname{ctr}} \to \N$ is unique if it exists
\end{enumerate}
\end{prop}

\begin{remark} The monoid $Q^{\operatorname{GS}}$ depends on a choice of \emph{type}, in this case a choice of the $u_q$ (the $u_p$ are already determined by the choice of $\alpha$). Here we are not allowed to choose a random type, but have to choose one which is compatible with the underlying stable map, in the sense that the modified tropical balancing condition is satisfied. The Gathmann-style condition (1) is a necessary condition for the existence of a compatible type (on the level of $\Pic C$).

Remember that we get one balancing condition for each generator of $\Gamma(C, f^* \overline{\Mcal}_H)$, so as well as there being a balancing condition for each internal component there is also one wherever the curve intersects the divisor in an isolate point (where for instance we get that $u_q = m_q$ at nodes connecting an external to an internal component). Also we require this condition to take place in $\Pic(C)$, but we are entitled at any point to restrict to components of $C$ or take degrees to get (in general weaker) conditions. \end{remark}

\begin{remark} Gathmann's condition obviously extends to the case $\Sigma \alpha \leq d$. \basenote{What is the correct generalisation of the compatibility condition?} Here we try to make the condition for a curve component of $f^{-1}(Y)$ into an easily verifiable criterion: if $Z$ is a smooth elliptic curve, recall that every line bundle of positive degree on it is effective, hence the relative condition can be rephrased as the following numerical criterion:
\begin{equation*} f_*[Z]\cdot [Y]+\sum_{j=1}^r m^{(j)}\geq \sum_{x_i\in Z}\alpha_i \end{equation*}
together with the additional requirement that, when the above is an equality, there is an isomorphism of line bundles:
\begin{equation}\label{eqn:eq_in_Pic}
(f|_{Z})^*\OO_X(Y) \otimes \OO_Z\left(\sum_{j=1}^r m^{(j)}y_j\right)=\OO_Z\left(\sum_{x_i\in Z}\alpha_ix_i\right) 
\end{equation}

If $Z$ is reducible, the correct extension is not to impose the same condition componentwise; rather, we should ask the numerical condition for the total degree, and, if $Z$ is obtained by gluing a smooth elliptic curve $E$ with a forest of rational trees $T_i$ at roots $q^E_i$, then the line bundle equality \eqref{eqn:eq_in_Pic} should be required in $\Pic(E)$ after counting each $q^E_i$ towards the left hand side with multiplicity:
\[f_*[T_i]\cdot [Y]+\sum_{\text{external components attached to $T_i$}} m^{(j)} \geq \sum_{x_j\in T_i}\alpha_j\]
%That this is the right thing to do follows from the proof that Gathmann's relative space is the projection of Jun Li's moduli space of maps to the expanded pair after contracting the higher levels of the accordion.
\end{remark}

\begin{proof}[Proof of Proposition \ref{closed pts of relative space}]
First suppose that the conditions (1) and (2) of the Proposition are satisfied; so we have Gathmann's condition on the level of line bundles and a factorisation:
\bcd
Q^{\operatorname{log}} \ar[r]\ar[d] & Q^{\operatorname{GS}} \ar[d] \\
Q^{\operatorname{ctr}}\ar[r] & \N
\ecd
We will construct a smoothing of the centrally aligned map, i.e. a $1$-parameter family whose central fibre is the map we started with and whose general fibre is in the nice locus. The construction proceeds in the following order:
\begin{enumerate}
\item construct the smoothing $\Ccal$ of the source curve $C$, as a scheme;
\item construct the extension $(\Mcal_{\Ccal},\delta)$ of the central alignment on $C$;
\item construct the extension $\Lcal$ of the line bundle $L=f^*\OO_{\PP^N}(H)$ from $C$ to $\Ccal$;
\item construct the extension of the section $u_0 \in \HH^0(C,L)$;
\item construct the extensions of the sections $u_1,\ldots,u_N$.
\end{enumerate}

We start with (1), i.e. the construction of the source curve $\Ccal$ as a family over $\spec \mathbbm{k}\llbracket t \rrbracket$. Let $E \subseteq C$ be the circuit, i.e. the minimal genus one subcurve. This is either a smooth elliptic curve or a wheel of rational curves. Look at the corresponding moduli point $[E] \in \Mfrak_{1,n}$. Since this is a smooth stack, we may choose an \'etale chart which locally looks like the tangent space of the stack, which in this case is the deformation space of the elliptic curve. This contains a factor $\Aaff^p$ where $p$ is the number of nodes of $C$; the $i$th co-ordinate axis corresponds to a smoothing of the $i$th node, keeping the others fixed, while the generic point of the corresponding orthogonal hyperplane corresponds to a deformation which smooths all the nodes except the $i$th one.

We need to choose a map $\formLine \to \Mfrak_{1,n}$ which sends the closed point to $[E]$ and the generic point to the locus of smooth curves. By the discussion of the previous paragraph, we can do this by choosing a map $\formLine \to \Aaff^p$ which sends the closed point to the origin and the generic point to the complmenet $\Gm^p \subseteq \Aaff^p$ of the co-ordinate axes. We can certainly choose such a map; furthermore, if for each node $q$ we choose a positive integer $r_q$, then we can choose the map so that it has tangency order $r_q$ to the hyperplane of $\Aaff^p$ corresponding to $q$. In terms of the resulting curve $\Ccal$ over $\formLine$, this means that if $C_1, C_2 \subseteq \Ccal$ are the two branches of $C$ around $q$, then these are $\QQ$-Cartier divisors in $\Ccal$, with:
\begin{equation*} C_1 \cdot C_2 = 1/r_q \end{equation*}
The weights $r_q$ are chosen by looking at the map $Q^{\operatorname{log}} \to \N$; this map is local since the map $Q^{\operatorname{log}} \to Q^{\operatorname{GS}}$ is local\basenote{Is this always true? Isn't this only true if there exists a log lifting? But maybe it's always true for Deligne--Faltings pairs.} and the map $Q^{\operatorname{GS}} \to \N$ is local by assumption. Thus each generator $e_q \in Q^{\operatorname{log}}$ maps to a positive integer $r_q$, and this is the weight we choose. We have thus constructed the smoothing $\Ccal$ of $C$.

We now move on to step (2), which involves extending the central alignment from $C$ to $\Ccal$. This is a relative trivial step: since the general fibre of $\Ccal$ is smooth, there is no choice of ordering of the markings or of contraction radius, and its minimal centrally aligned monoid is trivial. We can thus take the log structures on the central fibre as charts for the log structures on the deformation. To be more precise, in the base we have a chart:
\begin{align*} Q^{\operatorname{ctr}} & \to \mathbbm{k} \llbracket t \rrbracket \\
e_q & \mapsto t^{r_q} \end{align*}
Note that this is well-defined by the fact that we have the factorisation $Q^{\operatorname{ctr}} \to \N$. This defines a log structure $\Mcal_S^{\operatorname{ctr}}$ on $S$. On the other hand the morphism $S \to \Mfrak_{1,n}$ defines another log structure $\Mcal_S^{\operatorname{log}}$ on $S$ corresponding to the minimal structure of a log smooth curve on $\Ccal$. There is a map $\Mcal_S^{\operatorname{log}} \to \Mcal_S^{\operatorname{ctr}}$ induced by the map of charts $Q^{\operatorname{log}} \to Q^{\operatorname{ctr}}$, and this gives a diagram:
\bcd
& (C,\Mcal_C^{\operatorname{log}}) \ar[d] \\
(S,\Mcal_S^{\operatorname{ctr}}) \ar[r] & (S,\Mcal_S^{\operatorname{log}})
\ecd
We define $(\Ccal,\Mcal_{\Ccal}^{\operatorname{ctr}})$ to be the fibre product. This is clearly a log smooth centrally aligned curve over $(S,\Mcal_S^{\operatorname{ctr}})$, with radius $\delta$ given by the radius on the central fibre (note that $\overline{\Mcal_S}^{\operatorname{ctr}}$ is supported on the closed point). This completes step (2).

We now begin step (3), that of extending the line bundle $L=f^*\OO_{\PP^N}(H)$ on $C$ to a line bundle $\Lcal$ on $\Ccal$. In case it hasn't been mentioned already, we assume that $H=\{ z_0=0\}$ is a co-ordinate hyperplane.

Let us write $C$ in terms of its irreducible components as $C = C_1 \cup \ldots \cup C_m$ with $C_1,\ldots,C_r$ internal and the rest external. We define the line bundle $\Lcal$ as
\begin{equation*} \Lcal = \OO_{\Ccal}\left(\sum_{i=1}^r \gamma_i C_i + \sum_{i=1}^n \alpha_i \tilde{x}_i \right) \end{equation*}
where each $\gamma_i$ is a positive integer obtained as the image under $Q^{\operatorname{GS}} \to \N$ of the generator corresponding to $C_i$, and the $\tilde{x}_i$ are extensions of the Cartier divisors $x_i$ to all of $\Ccal$.

This comes with a natural section (or rather, a $\CC^*$-equivalence class of sections) $\tilde{u}_0$ given by the divisor in brackets above, and this can be chosen in a unique way so that it restricts to $u_0$ on the central fibre of $\Ccal$. This completes step (4).

Now we must verify that $\Lcal|_C=L$. For each external component $C_j$, we have
\begin{equation*} \Lcal|_{C_j} = \OO_{C_j} \left( \sum_{x_i \in C_j} \alpha_i x_i + \sum_{q_i \in C_j} (\gamma_i/r_i) q_j \right) \end{equation*}
where the latter sum is over nodes of $C_j$ which connect $C_j$ to an internal component $C_i$. We thus need
\begin{equation*} m_i = \gamma_i / r_i \end{equation*}
in order to conclude that $\Lcal|_{C_j} = L|_{C_j}$. But this is true, because in $Q^{\operatorname{GS}}$ the generator corresponding to the internal component $C_i$ is identified with $u_{q_i}$ times the generator corresponding to $q_i$, and $u_{q_i}=m_i$ by the fact that the type we chose to define $Q^{\operatorname{GS}}$ must satisfy the tropical balancing condition. Pushing forward this relation from $Q^{\operatorname{GS}}$ to $\N$ gives the equation above. This argument also applies when we have an external subcurve $Z \subseteq C$ which is a wheel of rational components.

Now let us consider an internal component $C_j$. We have
\begin{align*} \Lcal|_{C_j} & = \OO_{C_j} \left( \sum_{x_i \in C_j} \alpha_i x_i + \sum_{\tilde{q}_i \in C_j} (\gamma_i/r_i) \tilde{q}_i \right) \otimes \OO_{\mathcal{C}}\left(- \gamma_j \sum_{i \neq j} C_i \right) \\
& = \OO_{C_j}\left( \sum_{x_i \in C_j} \alpha_i x_i + \sum_{q_i \in C_j} \left(\frac{\gamma_i - \gamma_j}{r_i}\right) q_i - \sum_{\tilde{q}_i \in C_j} m_i \tilde{q}_i \right)\end{align*}
where the middle sum is over adjacent internal components and the final sum is over adjacent external components. The fact that $\gamma_j/r_i = m_i$ follows as in the previous case.

We must argue that $\Lcal|_{C_j} = L|_{C_j}$. Suppose for the moment that $C_j$ is genus zero, so that we only need to check the degrees. We must show:
\begin{equation*} \deg L|_{C_j} = \sum_{x_i \in C_j} \alpha_i - \sum_{q_i \in C_j} \left( \dfrac{\gamma_j - \gamma_i}{r_i}\right) - \sum_{\tilde{q}_i \in C_j} u_{\tilde{q}_i}\end{equation*}
But by the definition of $Q^{\operatorname{GS}}$ we have:
\begin{equation*} (\gamma_j - \gamma_i)/r_i = u_{q_i} \end{equation*}
From which the above formula holds because of the tropical balancing condition we imposed when choosing the type.

Similarly if $C_j$ is smooth of genus one we must show:
\begin{equation*} L|_{C_j} = \OO_{C_j} \left( \sum_{x_i \in C_j} \alpha_i x_i - \sum_{q_i \in C_j} u_{q_i} q_i - \sum_{\tilde{q}_i \in C_j} u_{\tilde{q}_i} \tilde{q}_i \right) \end{equation*}
But this is just the balancing condition on the level of $\Pic(C_j)$.

What about if $C_j$ is a wheel of rational components? Then it isn't irreducible, so let's write it as $Z= C_1 \cup \ldots \cup C_k$. So there are other internal components $C_{k+1},\ldots,C_r$ and external components $C_{r+1},\ldots,C_m$. We assume that $C_1,\ldots,C_k$ are ordered so that $C_i$ intersects $C_{i-1}$, $C_{i+1}$ and none others. Notice that this means that:
\begin{equation*} C_i^2 = -1/r_{i-1} - 1/r_{i+1} - \sum_{q_j \in C_i} 1/r_j \end{equation*}
for all $i \in \{1,\ldots,k\}$, where the sum is over all nodes $q_j$ which connect $Z$ to the rest of the curve. Thus we obtain:
\begin{align*} \Lcal|_Z & = \OO_Z \left( \sum_{x_i \in Z} \alpha_i x_i\right) \otimes \OO_{\Ccal} \left( \sum_{i=1}^r \gamma_i C_i \right)\bigg|_Z \\
& = \OO_Z \left( \sum_{x_i \in Z} \alpha_i x_i + \sum_{\substack{q_i \in Z \\ \text{$q_i$ joins $Z$} \\ \text{to int. cpt.}}} (\gamma_i/r_i) q_i \right) \otimes \OO_{\Ccal} \left( \sum_{i=1}^k \gamma_i C_i \right) \bigg|_Z \\
& = \OO_Z \left( \sum_{x_i \in Z} \alpha_i x_i + \sum_{\substack{q_i \in Z \\ \text{$q_i$ joins $Z$} \\ \text{to int. cpt.}}} (\gamma_i/r_i) q_i + \sum_{i=1}^k \gamma_i \left( \frac{1}{r_{i,i-1}} q_{i-1} + \frac{1}{r_{i,i+1}} q_{i+1} + (C_i^2) s_i \right)\right)
\end{align*}
where $q_{i-1}$ is the node connecting $C_{i-1}$ and $C_i$ (and similarly for $q_{i+1}$) and $s_i$ is some point of $C_i$ (doesn't really matter which, since $C_i$ is rational).

Using the description of $C_i^2$ given above, we see that the terms $q_{i-1}/r_{i,i-1}$ and $q_{i+1}/r_{i,i+1}$ cancel, and we are left with
\begin{align*} \Lcal|_Z & = \OO_Z \left( \sum_{x_i \in Z} \alpha_i x_i + \sum_{\substack{q_i \in Z \\ \text{$q_i$ joins $Z$} \\ \text{to int. cpt.}}} (\gamma_i/r_i) q_i + \sum_{\substack{q_i \in Z \\ \text{$q_i$ joins $Z$} \\ \text{to rest of curve}}} (\gamma_i/r_i) q_i \right) \\
& = \OO_Z \left( \sum_{x_i \in Z} \alpha_i x_i + \sum_{\substack{q_i \in Z \\ \text{$q_i$ joins $Z$} \\ \text{to int. cpt.}}} \left( \dfrac{\gamma_i-\gamma_j}{r_i} \right) q_i - \sum_{\substack{q_i \in Z \\ \text{$q_i$ joins $Z$} \\ \text{to ext. cpt}}} (\gamma_i/r_i) q_i \right)
\end{align*}
and this is precisely the balancing condition on the level of $\Pic(Z)$. This proves that $\Lcal|_C = L$, except for one case which we haven't yet considered, namely when there is a wheel of rational curves which is partially internal and partially external. In this case essentially the same argument as given above applies.

We have thus shown that we can break $C$ up into pieces (mostly into its irreducible components, unless there is a rational wheel in which case we treat this as a single piece) such that the corresponding intersection graph has genus zero and such that $\Lcal$ restricts to $L$ on each of the pieces. The fact that $\Lcal|_C = L$ then immediately follows (since the intersection graph has genus zero, so the Picard group of $C$ is equal to the direct sum of the Picard groups of the pieces). This completes the proof of step (3).
